## Load Dataset 

```{r}
library(readr)
library(dplyr)
International_Report_Departures <- read_csv("International_Report_Departures.csv")
International_Report_Departures

International_Report_Passengers <- read_csv("International_Report_Passengers.csv")
International_Report_Passengers
```

## Data Exploration
```{r}
summary(International_Report_Departures)
summary(International_Report_Passengers)
```

## Data Cleaning
```{r}
check_missing_values <- function(data) {
  missing_values <- sapply(data, function(x) sum(is.na(x)))
  return(missing_values)
}
check_missing_values(International_Report_Departures)
check_missing_values(International_Report_Passengers)
```

## Check for outliers in the dataset
```{r}
library(ggplot2)
# Assuming 'data' is your dataset and 'attribute' is the column you want to check for outliers
check_outliers <- function(data, attribute) {
  ggplot(data, aes(y = .data[[attribute]])) +
    geom_boxplot() +
    labs(title = paste("Box Plot for", attribute), y = attribute) +
    theme_minimal()
}
```
```{r}
# display boxplot for a specific attribute, replace 'attribute_name' with the actual column name
check_outliers(International_Report_Departures, "Total")
```
# Boxplot for both Departures and Passengers they can both be refined to show specific years(for major events 9/11, 2008 recesion) 
```{r}
filtered_data <- subset(International_Report_Departures, Year >= 2000 & Year <= 2011)
Pass_data <- subset(International_Report_Passengers, Year >= 2000 & Year <= 2011)
boxplot(Total ~ Year,
         data = International_Report_Departures,
         main = "Departures by Airline",
         xlab = "Year",
         ylab = "Total Departures",
         las = 1)
boxplot(Total ~ Year,
         data = International_Report_Passengers,
         main = "Departures by Airline",
         xlab = "Year",
         ylab = "Total Departures",
         las = 1)
```
## Total Departures by Year or Month

```{r}
departures_by_year <- International_Report_Departures %>%
  group_by(Year) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_year
```

```{r}
departures_by_month <- International_Report_Departures %>%
  group_by(Year, Month) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_month
```

## Departures by Airport

```{r}
departures_by_origin <- International_Report_Departures %>%
  group_by(usg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_origin
```

```{r}
departures_by_dest <- International_Report_Departures %>%
  group_by(fg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_dest
```

## Departures by Airline

```{r}
departures_by_airline <- International_Report_Departures %>%
  group_by(carrier) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_airline
```

## Route Analysis

```{r}
departures_by_route <- International_Report_Departures %>%
  group_by(usg_apt, fg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_route
```

## Seasonal Patterns

```{r}
seasonal_pattern <- International_Report_Departures %>%
  group_by(Month) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
seasonal_pattern
```

## Top Routes or Airports

```{r}
top_n_routes <- departures_by_route %>%
  arrange(desc(TotalDepartures)) %>%
  slice_head(n = 10)
top_n_routes
```

```{r}
top_n_airports <- departures_by_origin %>%
  arrange(desc(TotalDepartures)) %>%
  slice_head(n = 10)
top_n_airports
```

## Year-over-Year Growth

```{r}
departures_by_year <- departures_by_year %>%
  arrange(Year) %>%
  mutate(YoY_Growth = (TotalDepartures - lag(TotalDepartures)) / lag(TotalDepartures) * 100)
departures_by_year
```

## Average Departures per Route/Airport

```{r}
avg_departures_per_route <- departures_by_route %>%
  summarise(AvgDepartures = mean(TotalDepartures, na.rm = TRUE))
avg_departures_per_route
```

```{r}
avg_departures_per_airport <- departures_by_origin %>%
  summarise(AvgDepartures = mean(TotalDepartures, na.rm = TRUE))
avg_departures_per_airport
```

## Visualization


## Visualization

### Total Departures by Year

```{r}
ggplot(departures_by_year, aes(x = Year, y = TotalDepartures)) +
  geom_line() +
  labs(title = "Total Departures by Year", x = "Year", y = "Total Departures") +
  theme_minimal()
```

### Top 10 Origin Airports

```{r}
ggplot(top_n_airports, aes(x = reorder(usg_apt, -TotalDepartures), y = TotalDepartures)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Origin Airports by Departures", x = "Origin Airport", y = "Total Departures") +
  theme_minimal()
```

### Departures Heatmap by Month and Year

```{r}
ggplot(departures_by_month, aes(x = factor(Month), y = factor(Year), fill = TotalDepartures)) +
  geom_tile() +
  labs(title = "Departures Heatmap by Month and Year", x = "Month", y = "Year") +
  scale_fill_viridis_c() +
  theme_minimal()
```
