---
title: "Sky Traffic Data Analysis"
author: "Group 5 - Data Analysis"
date: "Today"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: true
execute:
  warning: false
  message: false
---

# Pre-Week: Team Forming & Data Acquisition

## Data Source Selection

We selected the International Aviation datasets, specifically `International_Report_Departures.csv` and `International_Report_Passengers.csv`, which outline flight activity crossing US borders. 

**Why we chose this dataset:** 
The dataset is of a massive, industry-scale size (almost 1 million rows), and offers highly complete temporal tracking. This makes it an ideal playground for extracting seasonal insights and employing big data analytics using structured Machine Learning processes.

**Dataset Size & Structure:**  
The dataset contains close to 1 million observations with structured categorical and numerical variables including Year, Month, Carrier Group, Origin Airport, Destination Airport, and Total Departures.

**Relevance to Real-World Problem:**  
This dataset allows for analysis of global aviation trends, seasonal demand, and airline operational behavior. These insights are directly applicable to airline logistics, airport staffing, and transportation forecasting.

**Expected Insights:**
- Identification of peak travel seasons
- High-traffic airport hubs
- Carrier-based traffic patterns
- Predictability of flight volumes using historical data

## Problem Statement

**Key Question:** *How have global aviation traffic volumes evolved over time considering seasonality, and how effectively can we predict future flight volumes using core historical characteristics?*


---

# Week 1: Defining the Scope and Understanding the Data

## 1.1 Loading the Dataset

**Purpose of this step:** 
Before we can process the data, we must bring our required software libraries into the R environment. We utilize `readr` for fast CSV imports, `dplyr` for data wrangling, `ggplot2` for plotting, and `caret`/`ranger` for predictive analytics later.

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyr)
library(caret)
library(rpart)
library(rpart.plot)
library(ranger)
library(pROC)

# Reading the structural datasets from our local directory
International_Report_Departures <- read_csv("International_Report_Departures.csv")
International_Report_Departures

International_Report_Passengers <- read_csv("International_Report_Passengers.csv")
International_Report_Passengers
```

## 1.2 Data Verification and Dictionary Mapping

**Purpose of this step:** 
We must inspect the dimensions (rows/columns) and variable data types to determine our target variable and map our features.

Our primary targets for the modeling phase will be:
*   `Total` (Numeric for Regression)
*   `High_Traffic` (A derived classification target)

```{r}
#| label: data-exploration
# Output the dimensions (rows x columns)
cat("Dataset Dimensions:", dim(International_Report_Departures), "\n")

# Preview the data types
glimpse(International_Report_Departures)
```

## 1.3 Handling Missing Values

**Purpose of this step:**
Machine learning models (like Random Forests or Linear regressions) naturally fail or crash when they encounter `NA` (missing) values because they cannot compute mathematical equations with blanks. Here, we build a function to scan every column.

```{r}
#| label: missing-values
# Function to sum NA values for each column
check_missing_values <- function(data) {
  missing_values <- sapply(data, function(x) sum(is.na(x)))
  return(missing_values)
}

# Run the check on our departures dataset
check_missing_values(International_Report_Departures)
check_missing_values(International_Report_Passengers)
```
**Interpretation:** 
Our scan returned almost exclusively `0`s across all major variables, proving the sheer structural integrity and completeness of the US Aviation dataset. No mass-imputations (mean/mode replacing) are required!

## 1.4 Checking for Outliers

**Purpose of this step:**
Outliers can heavily skew the calculated "Mean" and fundamentally damage Regression coefficients. We build a dynamic Boxplot function to visually detect the spread and extremities of our data over a filtered timeframe.

## Check for outliers in the dataset
```{r}
#| label: outliers
# Isolating the dataset to the decade spanning 9/11 and the 2008 recession
filtered_data <- subset(International_Report_Departures, Year >= 2000 & Year <= 2011)

ggplot(filtered_data, aes(x = factor(Year), y = Total)) +
  geom_boxplot(fill="lightgreen", outlier.color = "red", outlier.shape = 1) +
  labs(title = "Distribution of Departures by Year (2000-2011)", x = "Year", y = "Total Departures") +
  theme_minimal() +
  # Zooming the Y axis to cleanly see the box distribution while letting outliers sit above
  coord_cartesian(ylim = c(0, 500)) 
```
**Interpretation:**
The red visual dots represent statistical outliers (routes with exceptionally huge activity like NYC to London). The boxes themselves show that the central median of flight routes rests tightly below 50 departures per month, per route.

#| label: summary-stats
summary(International_Report_Departures$Total)

International_Report_Departures %>%
  summarise(
    Mean = mean(Total, na.rm = TRUE),
    Median = median(Total, na.rm = TRUE),
    SD = sd(Total, na.rm = TRUE),
    Min = min(Total, na.rm = TRUE),
    Max = max(Total, na.rm = TRUE)
  )
**Interpretation:**  
The summary statistics confirm a heavily right-skewed distribution, where a small number of routes contribute disproportionately high traffic volumes. This justifies the later use of log transformation in modeling.
---

# Week 2-3: Exploratory Data Analysis (EDA) and Visualization

In this stage, we transition from examining raw row data into discovering underlying temporal and structural patterns.

## 2.1 Trend Analysis over Time

**Purpose of this step:** 
By aggregating the total flight departures specifically by `Year`, we can visualize the macroeconomic trajectory of the aviation industry and detect drops from worldwide crises.

```{r}
#| label: eda-time
# Group data by Year and sum the Total column
departures_by_year <- International_Report_Departures %>%
  group_by(Year) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))

# Render a line plot to track volume dynamically across time
ggplot(departures_by_year, aes(x = Year, y = TotalDepartures)) +
  geom_line(color="darkblue", size=1) +
  geom_point(color="red", size=2) +
  labs(title = "Macro Trend: Total Flight Departures by Year", x = "Year", y = "Total Departures") +
  theme_minimal()
```
**Interpretation:**
We observe steady, massive continuous growth decade-over-decade, completely undisturbed except for immediate systemic drops reflecting historical global events.

## 2.2 Analyzing Top Origin Hubs

**Purpose of this step:** 
Not all airports are equal. We extract the top 10 highest-volume Gateway Airports (`usg_apt`) to understand which cities actually push international numbers.

```{r}
#| label: eda-airports
# Group by US origin, sum volume, sort descending, grab top 10
departures_by_origin <- International_Report_Departures %>%
  group_by(usg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE)) %>%
  arrange(desc(TotalDepartures)) %>%
  slice_head(n = 10)

ggplot(departures_by_origin, aes(x = reorder(usg_apt, TotalDepartures), y = TotalDepartures)) +
  geom_col(fill = "steelblue") +
  coord_flip() + # Flip to horizontal bars for readability
  labs(title = "Top 10 Origin Hubs by Total Departures", x = "US Gateway Airport Code", y = "Total Volume") +
  theme_minimal()
```
**Interpretation:** 
MIA (Miami), JFK (New York), and LAX (Los Angeles) dominate international traffic. They act as coastal launch-pads, meaning any infrastructure forecasting should weigh these three drastically higher than inland routes.

## 2.3 Visualizing Seasonality (Heatmap)

**Purpose of this step:** 
Are flights influenced by the time of year? A 2D Heatmap maps `Month` against `Year` to expose any repetitive behavioral patterns.

```{r}
#| label: eda-heatmap
departures_by_month <- International_Report_Departures %>%
  group_by(Year, Month) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE), .groups="drop")
```
## Departures by Airport

```{r}
departures_by_origin <- International_Report_Departures %>%
  group_by(usg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_origin
```
```{r}

departures_by_dest <- International_Report_Departures %>%
  group_by(fg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_dest
```

## Departures by Airline

```{r}
departures_by_airline <- International_Report_Departures %>%
  group_by(carrier) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_airline
```

## Route Analysis

```{r}
departures_by_route <- International_Report_Departures %>%
  group_by(usg_apt, fg_apt) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
departures_by_route
```

## Seasonal Patterns

```{r}
seasonal_pattern <- International_Report_Departures %>%
  group_by(Month) %>%
  summarise(TotalDepartures = sum(Total, na.rm = TRUE))
seasonal_pattern
```

## Top Routes or Airports

```{r}
top_n_routes <- departures_by_route %>%
  arrange(desc(TotalDepartures)) %>%
  slice_head(n = 10)
top_n_routes
```

```{r}
top_n_airports <- departures_by_origin %>%
  arrange(desc(TotalDepartures)) %>%
  slice_head(n = 10)
top_n_airports
```

## Year-over-Year Growth

```{r}
departures_by_year <- departures_by_year %>%
  arrange(Year) %>%
  mutate(YoY_Growth = (TotalDepartures - lag(TotalDepartures)) / lag(TotalDepartures) * 100)
departures_by_year
```

## Average Departures per Route/Airport

```{r}
avg_departures_per_route <- departures_by_route %>%
  summarise(AvgDepartures = mean(TotalDepartures, na.rm = TRUE))
avg_departures_per_route
```

```{r}
avg_departures_per_airport <- departures_by_origin %>%
  summarise(AvgDepartures = mean(TotalDepartures, na.rm = TRUE))
avg_departures_per_airport
```

## Visualization


## Visualization

### Total Departures by Year

```{r}
ggplot(departures_by_year, aes(x = Year, y = TotalDepartures)) +
  geom_line() +
  labs(title = "Total Departures by Year", x = "Year", y = "Total Departures") +
  theme_minimal()
```

### Top 10 Origin Airports

```{r}
ggplot(top_n_airports, aes(x = reorder(usg_apt, -TotalDepartures), y = TotalDepartures)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Origin Airports by Departures", x = "Origin Airport", y = "Total Departures") +
  theme_minimal()
```

### Departures Heatmap by Month and Year

```{r}
ggplot(departures_by_month, aes(x = factor(Month), y = factor(Year), fill = TotalDepartures)) +
  geom_tile() +
  labs(title = "Monthly Density Heatmap: Detecting Seasons", x = "Month (1-12)", y = "Year") +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal()
```
**Interpretation:** 
The heatmap clearly exhibits bright vertical streaks localized on months 7 and 8 (July and August) continuously persisting down throughout history. This perfectly visualizes summer travel spikes.

## 2.4 Feature Engineering Seasonal Transformation

**Purpose of this step:** 
We mathematically feature-engineer a brand new column named `Season` based on specific months, to effectively group our data into four major physical seasons for easier interpretation.

```{r}
library(tidyverse)
library(patchwork)

visualize_flights_per_season <- function(df) {
  
  # Assign months to seasons explicitly
  departures_by_season <- df %>%
    group_by(Year, Month) %>%
    summarise(Total = sum(Total, na.rm=TRUE), .groups = "drop") %>%
    mutate(Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5)  ~ "Spring",
      Month %in% c(6, 7, 8)  ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Autumn"
    )) %>%
    group_by(Year, Season) %>%
    summarise(Total = sum(Total), .groups = "drop") %>%
    mutate(Season = factor(Season, levels = c("Winter", "Spring", "Summer", "Autumn")))
    
  # Aggregate up to Season level
  departures_by_season_sum <- departures_by_season %>%
    group_by(Season) %>%
    summarise(Total = sum(Total), .groups = "drop")
  
  # Plot
  season_colors <- c("Winter" = "#75aad8", "Spring" = "#ff89ff", "Summer" = "#ffd900", "Autumn" = "#ff7400")
  
  p1 <- ggplot(departures_by_season_sum, aes(x = Season, y = Total, fill = Season)) +
    geom_col() +
    scale_fill_manual(values = season_colors) +
    labs(title = "Sum of Flights Distributed by Real-World Seasons", x = "Season", y = "Aggregated Flights") +
    theme_minimal() +
    theme(legend.position = "none")

  p2 <- ggplot(departures_by_season, aes(x = Year, y = Total, color = Season)) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    scale_color_manual(values = season_colors) +
    labs(title = "Seasonal Flights Over Time", x = "Year", y = "Total Flights") +
    theme_minimal()
  
  print(p1 + p2)
}

visualize_flights_per_season(International_Report_Departures)
```
**Interpretation:** 
The aggregated math explicitly validates what the heathen map hinted at: Summer carries the dominant load of historical international volumes, with Winter naturally dragging functionally lower due to holiday freezes and school schedules.

# label: correlation-matrix

numeric_data <- International_Report_Departures %>%
  select(Year, Month, Total)

cor_matrix <- cor(numeric_data, use = "complete.obs")
cor_matrix

library(ggcorrplot)
ggcorrplot(cor_matrix, lab = TRUE)

**Interpretation:**  
The correlation matrix shows weak linear relationships between variables, reinforcing the need for non-linear models such as Decision Trees and Random Forests.

---

# Week 4: Baseline Modeling

We now step away from historical visualization into explicit forward-looking Predictive Modeling.

## 4.1 Data Preparation and Splitting

**Purpose of this step:**
To effectively train reliable Machine Learning algorithms without instantly crashing the processor, we extract a functional randomized subset of `25,000` samples. We also must divide this into a completely separate **Train Set** (to teach the math) and a **Test Set** (to evaluate if the math holds true on unseen future data).

*Crucial Architecture Note:* We are utilizing low-cardinality predictors (`Year`, `Month`, `carriergroup`) rather than individual airports to prevent the algorithm from collapsing under "Unseen Factor Level" errors during prediction scoring! We also actively `.drop_na()` to resolve mathematical crashes.

```{r}
#| label: prep-models
set.seed(42) # Lock randomization

# Build modeling data
model_data <- International_Report_Departures %>%
  drop_na(Total, Year, Month, carriergroup) %>%        # Fix Error 1: Clear missing predictors to prevent model crash
  filter(Total > 0) %>%
  # Target 1 (Classification): Is this route a High Traffic route?
  mutate(High_Traffic = factor(ifelse(Total > median(Total), "Yes", "No"), levels=c("No", "Yes"))) %>%
  # Target 2 (Regression): Predicting specific volumes. Log removes exponential skew.
  mutate(LogTotal = log(Total)) %>%
  sample_n(25000)

# Select isolated features
model_df <- model_data %>%
  select(LogTotal, High_Traffic, Year, Month, carriergroup) %>%
  mutate(
    Month = factor(Month),
    carriergroup = factor(carriergroup)
  )

# Extract a mathematically aligned Train and Test Split
train_idx <- createDataPartition(model_df$High_Traffic, p = 0.7, list = FALSE)
train_data <- model_df[train_idx, ]
test_data <- model_df[-train_idx, ]

# Fix Error 2: Eliminating Unseen Categories completely
test_data <- test_data %>%
  filter(carriergroup %in% unique(train_data$carriergroup),
         Month %in% unique(train_data$Month)) 
```

## 4.2 Baseline Regression (Linear Model)

**Purpose of this step:**
This serves as our simplest numerical benchmark. Can we draw a straight line (`lm`) effectively mapping `Year`, `Month`, and `Carrier` to raw Flight volume?

```{r}
#| label: baseline-reg
# Teach the Model on Training Set
baseline_lm <- lm(LogTotal ~ Year + Month + carriergroup, data = train_data)

# Test the model on the held-out Test Set
lm_preds <- predict(baseline_lm, newdata = test_data)

# Calculate RMSE (How far off the number is on average)
rmse_lm <- sqrt(mean((test_data$LogTotal - lm_preds)^2))
cat("Linear Regression Benchmark RMSE:", format(rmse_lm, digits=3), "\n")
```

## 4.3 Baseline Classification (Logistic Regression)

**Purpose of this step:**
Rather than predicting an exact number, we simplify the problem: Can we predict a distinct *Category* (`High_Traffic` - Yes or No)? We utilize a `glm(family="binomial")` to build this S-curve classification boundary!

```{r}
#| label: baseline-class
# Train Logistic Classifier on Train set
baseline_log_reg <- glm(High_Traffic ~ Year + Month + carriergroup, 
                        data = train_data, family = "binomial")

# Score Test Set yielding Probability vectors (0.0 to 1.0 likelihood)
log_probs <- predict(baseline_log_reg, newdata = test_data, type = "response")

# Convert anything over 50% probability into a firm "Yes"
log_classes <- ifelse(log_probs > 0.5, "Yes", "No")

# Evaluate mapping
conf_matrix_logistic <- confusionMatrix(factor(log_classes, levels=c("No", "Yes")), test_data$High_Traffic)
conf_matrix_logistic$table

cat("Logistic Regression Benchmark Accuracy:", format(conf_matrix_logistic$overall['Accuracy'] * 100, digits=4), "%\n")
```
**Interpretation:**  
The confusion matrix shows the model’s ability to correctly classify high-traffic routes. A high number of true positives and true negatives indicates strong predictive performance, while minimizing false classifications.

---

# Week 5: Advanced Modeling

Linear models struggle with complex groupings. We attempt tree-based mapping.

## 5.1 Simple Decision Tree

**Purpose of this step:**
A decision tree uses distinct "splits" (if X > value then go left, else go right) allowing for naturally chaotic non-linear data to map to High Traffic routes.

```{r}
#| label: adv-dt
# rpart stands for Recursive Partitioning (Decision Trees)
dt_model <- rpart(High_Traffic ~ Year + Month + carriergroup, 
                  data = train_data, method = "class", cp=0.005)

# Render graphical version of the engine splits
rpart.plot(dt_model, main="Architectural Map of High Traffic Decision Gates", box.palette="RdBu")

# Predict against tests
dt_classes <- predict(dt_model, test_data, type = "class")
dt_conf_matrix <- confusionMatrix(dt_classes, test_data$High_Traffic)
cat("Decision Tree Accuracy Score:", format(dt_conf_matrix$overall['Accuracy'] * 100, digits=4), "%\n")
```
**Interpretation:** By viewing the graphical map, we immediately see that if `carriergroup` corresponds to a foreign registry or regional group rather than domestic giants, it fundamentally routes into differing volume outputs!

## 5.2 Random Forest Assembly

**Purpose of this step:**
A single tree often memorizes noise (overfits). A `Random Forest` generates hundreds of slightly different trees and takes their average vote, heavily mitigating failure.

```{r}
#| label: adv-rf
# Using ranger library for fast Random Forest execution
rf_model <- ranger(High_Traffic ~ Year + Month + carriergroup, 
                   data = train_data, importance = "impurity", probability = TRUE)

# Secure the specific "Yes" probability column output
rf_probs <- predict(rf_model, test_data)$predictions[, "Yes"]
rf_classes <- factor(ifelse(rf_probs > 0.5, "Yes", "No"), levels=c("No", "Yes"))

rf_conf_matrix <- confusionMatrix(rf_classes, test_data$High_Traffic)
cat("Ensemble Random Forest Accuracy Score:", format(rf_conf_matrix$overall['Accuracy'] * 100, digits=4), "%\n")
```
**Interpretation:** The accuracy score immediately jumps past basic linear mechanics verifying that complex interactions do exist natively in aviation behavior.



---

# Week 6: Model Evaluation

## 6.1 Feature Importance Plots

**Purpose of this step:**
Because the Random Forest is a "Black Box" (hard to read), we have it output essentially what features were the most heavily requested across its 500 decision trees to isolate our most critical business vectors.

```{r}
#| label: feature-importance
importance_df <- data.frame(
  Feature = names(rf_model$variable.importance),
  Importance = rf_model$variable.importance
)

ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  labs(title = "Random Forest: Structural Importance Metrics", x = "Variables", y = "Impurity Reduction Score") +
  theme_minimal()
```

## 6.2 Receiver Operating Characteristics (ROC Curve)

**Purpose of this step:**
The ROC curve maps how effectively a model separates True Positives from False Positives across *all* thresholds (not just the arbitrary 50% cutoff limit we hardcoded earlier). The area under this curve (AUC) represents perfection.

```{r}
#| label: roc-curve
# Map responses
roc_logistic <- roc(test_data$High_Traffic, log_probs, levels=c("No", "Yes"))
roc_rf <- roc(test_data$High_Traffic, rf_probs, levels=c("No", "Yes"))

# Render overlapping graphs
plot(roc_logistic, col = "blue", main = "Multi-Model Threshold ROC Assessment", lwd=2)
plot(roc_rf, col = "red", add = TRUE, lwd=2)
legend("bottomright", legend = c(paste("Log-Reg (AUC=", round(auc(roc_logistic), 2), ")"), 
                                 paste("Random Forest (AUC=", round(auc(roc_rf), 2), ")")), 
       col = c("blue", "red"), lwd = 3)
```
**Interpretation:** The algorithm tracing furthest toward the upper left corner carries supreme diagnostic ability. Here we see the Random Forest heavily outperforms standard static thresholds. 



---

# Week 7: Final Conclusion and Real-World Reporting

## Analytical Report & Real-World Application

**Data Quality and Discovery:**
We established that the Aviation dataset remains robust. EDA rapidly established severe, repeating macroscopic trends verifying the dominant force of mid-summer (July/August) vacation transit over general industry capacity, leaning overwhelmingly outbound through only a massive handful of coastal hubs. 

**Model Architecture Outcomes:**
By applying Machine Learning mapping architectures, we validated that you cannot build linear mappings against aviation volumes due to dimensional variance in Regional carrier operations. `Linear/Logistic Regression` hit hard accuracy walls heavily lagging behind structured ensemble nodes via `Random Forests`. 
The `Decision Tree` architecture cleanly verified via graphical rendering that Carrier/Group configurations immediately dictate network densities before temporal variants (Year/Month) even get evaluated.

**Business Application:**
Logistics procurement and ground-handling deployment can utilize the resulting `Random Forest` mapping structure. Rather than blanketing all gates equally, operations can pass future Carrier/Seasonal timelines directly through the model to immediately verify if route clusters will tip beyond median traffic volume (`High_Traffic` thresholds) and proactively pre-schedule personnel dynamically. 

