## Load Dataset 

```{r}

library(readr)
International_Report_Departures <- read_csv("International_Report_Departures.csv")
International_Report_Departures

International_Report_Passengers <- read_csv("International_Report_Passengers.csv")
International_Report_Passengers
```

## Data Exploration
```{r}
summary(International_Report_Departures)
summary(International_Report_Passengers)
```

## Data Cleaning
```{r}
check_missing_values <- function(data) {
  missing_values <- sapply(data, function(x) sum(is.na(x)))
  return(missing_values)
}
check_missing_values(International_Report_Departures)
check_missing_values(International_Report_Passengers)
```

## Check for outliers in the dataset
```{r}
library(ggplot2)
# Assuming 'data' is your dataset and 'attribute' is the column you want to check for outliers
check_outliers <- function(data, attribute) {
  ggplot(data, aes(y = .data[[attribute]])) +
    geom_boxplot() +
    labs(title = paste("Box Plot for", attribute), y = attribute) +
    theme_minimal()
}

# display boxplot for a specific attribute, replace 'attribute_name' with the actual column name
check_outliers(International_Report_Departures, "Total")
```
# Boxplot for both Departures and Passengers they can both be refined to show specific years(for major events 9/11, 2008 recesion) 
filtered_data <- subset(International_Report_Departures, Year >= 2000 & Year <= 2011)
Pass_data <- subset(International_Report_Passengers, Year >= 2000 & Year <= 2011)
boxplot(Total ~ Year,
+         data = International_Report_Departures,
+         main = "Departures by Airline",
+         xlab = "Year",
+         ylab = "Total Departures",
+         las = 1)
boxplot(Total ~ Year,
+         data = International_Report_Passengers,
+         main = "Departures by Airline",
+         xlab = "Year",
+         ylab = "Total Departures",
+         las = 1)

